---
title: "ANT4150 Final Project"
author: "Laura Hu"
date: "2024-05-05"
output:
  pdf_document: default
  html_document: default
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r}
#Load packages 
library(plyr)
library(dplyr)
library(ggplot2)

#import libraries 
library(readr)
master_data <- read_csv("master_anon_ANT4150.csv",
 col_types = cols_only(
 sex = col_guess(),
 period = col_guess(), 
 age = col_guess(), 
 grave.goods = col_guess(), 
 location = col_guess(),  
 burial = col_guess(),
 l.vo = col_guess(),
 l.voa = col_guess(),
 Zp.left = col_guess(),
 Zp.right = col_guess()),
 na = "NA")

#View(master_data)

master_data$sex <- as.factor(master_data$sex)
master_data$age <- as.factor(master_data$age)
master_data$period <- as.factor(master_data$period)
master_data$location <- as.factor(master_data$location)
master_data$burial <- as.factor(master_data$burial)

```

## Preliminary Work

-   feature engineering

-   normality check

```{r}

data_nona_vo_voa <- master_data[!is.na(master_data$l.vo) & !is.na(master_data$l.voa),]

ggplot(data_nona_vo_voa, aes(x = Zp.left, y = Zp.right)) + geom_point()
# Based on the graph, we can say that Zp.left and right about the same 
## Do we need to run t-test to see??


# Set a new variable that combines the average of left and right 
data_nona_vo_voa <- data_nona_vo_voa %>% mutate(avg_Zp = (Zp.left + Zp.right)/2)

for ( i in 1:nrow(data_nona_vo_voa)){
  if (is.na(data_nona_vo_voa$Zp.left[i]) == TRUE & is.na(data_nona_vo_voa$Zp.right[i]) == FALSE){
  data_nona_vo_voa$avg_Zp[i] = data_nona_vo_voa$Zp.right[i]
  } else if (is.na(data_nona_vo_voa$Zp.left[i]) == FALSE & is.na(data_nona_vo_voa$Zp.right[i]) == TRUE){
  data_nona_vo_voa$avg_Zp[i] = data_nona_vo_voa$Zp.left[i]
  }
  
}

# delete all data that are not avaliable for both femurs 
data_nona_vo_voa <- data_nona_vo_voa[!is.na(data_nona_vo_voa$avg_Zp),]


# Normality Check 
ggplot(data_nona_vo_voa, aes(sample = avg_Zp)) + stat_qq() + stat_qq_line() # Average Zp score is relatively normally distributed
ggplot(data_nona_vo_voa, aes(sample = l.vo)) + stat_qq() + stat_qq_line() # The reason why the graphs look like this is because they are in percentages and therefore is more discrete.
ggplot(data_nona_vo_voa, aes(sample = l.voa)) + stat_qq() + stat_qq_line()


# See the distribution of l.vo and l.voa
ggplot(data_nona_vo_voa, aes(x = avg_Zp)) + geom_histogram()
ggplot(data_nona_vo_voa, aes(x = l.vo)) + geom_histogram()
ggplot(data_nona_vo_voa, aes(x = l.voa)) + geom_histogram()

# See the relationship between l.vo and l.voa 
ggplot(data_nona_vo_voa, aes(x = l.voa, y = l.vo)) + geom_point()



```

## Initial Data Observation

### Average Zp Analysis

Average Zp score for each nominal variables (Vince has the analysis)

```{r}
ggplot(data_nona_vo_voa, aes(x = sex, y = avg_Zp, col = sex)) + geom_boxplot()
Zp_sex_summary <- ddply(data_nona_vo_voa, c("sex"), summarise, N = length(avg_Zp), mean = mean(avg_Zp), median = median(avg_Zp), sd = sd(avg_Zp))
Zp_sex_summary

ggplot(data_nona_vo_voa, aes(x = age, y = avg_Zp, col = age)) + geom_boxplot()
Zp_age_summary <- ddply(data_nona_vo_voa, c("age"), summarise, N = length(avg_Zp), mean = mean(avg_Zp), median = median(avg_Zp), sd = sd(avg_Zp))
Zp_age_summary

ggplot(data_nona_vo_voa, aes(x = location, y = avg_Zp, col = location)) + geom_boxplot()
Zp_location_summary <- ddply(data_nona_vo_voa, c("location"), summarise, N = length(avg_Zp), mean = mean(avg_Zp), median = median(avg_Zp), sd = sd(avg_Zp))
Zp_location_summary

ggplot(data_nona_vo_voa, aes(x = burial, y = avg_Zp, col = burial)) + geom_boxplot()
Zp_burial_summary <- ddply(data_nona_vo_voa, c("burial"), summarise, N = length(avg_Zp), mean = mean(avg_Zp), median = median(avg_Zp), sd = sd(avg_Zp))
Zp_burial_summary # test needed 

ggplot(data_nona_vo_voa, aes(x = period, y = avg_Zp, col = period)) + geom_boxplot()
Zp_period_summary <- ddply(data_nona_vo_voa, c("period"), summarise, N = length(avg_Zp), mean = mean(avg_Zp), median = median(avg_Zp), sd = sd(avg_Zp))
Zp_period_summary

ggplot(data_nona_vo_voa, aes(x = grave.goods, y = avg_Zp, col = grave.goods)) + geom_boxplot()
Zp_grave_summary <- ddply(data_nona_vo_voa, c("grave.goods"), summarise, N = length(avg_Zp), mean = mean(avg_Zp), median = median(avg_Zp), sd = sd(avg_Zp))
Zp_grave_summary # test needed 

```

### Lumbar Vo Analysis

```{r}
ggplot(data_nona_vo_voa, aes(x = l.vo, col = sex)) + geom_histogram() +
  facet_wrap(~sex)
lvo_sex_summary <- ddply(data_nona_vo_voa, c("sex"), summarise, N = length(l.vo), mean = mean(l.vo), median = median(l.vo), sd = sd(l.vo))
lvo_sex_summary
# It is hard to tell from the box plot, but female has a higher lumbar vo mean than males 

ggplot(data_nona_vo_voa, aes(x = l.vo, col = age)) + geom_histogram() +
  facet_wrap(~age)
lvo_age_summary <- ddply(data_nona_vo_voa, c("age"), summarise, N = length(l.vo), mean = mean(l.vo), median = median(l.vo), sd = sd(l.vo))
lvo_age_summary
# Boxplot is not the best visual for analysis, hence we want to use histogram to represent the visual. The average of the middle age group has about 12% average in lumbar%, young age group has no lumbar vo at all, older age group has about 12.5% avergae in lumbar vo%. 

ggplot(data_nona_vo_voa, aes(x = l.vo, col = location)) + geom_histogram() +
  facet_wrap(~location)
lvo_location_summary <- ddply(data_nona_vo_voa, c("location"), summarise, N = length(l.vo), mean = mean(l.vo), median = median(l.vo), sd = sd(l.vo))
lvo_location_summary
# Use histogram, people who were buried at church has no lumbar vo, people who buried at churchyard has about 8% average in lumbar vo, and people who were buried at monastic has about 18% average in lumbar vo. The average of monastic is skewed by that one person with 100% lumbar vo. 

ggplot(data_nona_vo_voa, aes(x = l.vo, col = burial)) + geom_histogram() +
  facet_wrap(~burial)
lvo_burial_summary <- ddply(data_nona_vo_voa, c("burial"), summarise, N = length(l.vo), mean = mean(l.vo), median = median(l.vo), sd = sd(l.vo))
lvo_burial_summary
# Use histogram, people who were buried earthen has about 7% lumbar vo, while people who buried at tomb has about 13% lumbar vo on average. People who are buried in tomb has more individuals with higher lumbar vo% than those who were buried earthen. 

ggplot(data_nona_vo_voa, aes(x = l.vo, col = period)) + geom_histogram() +
  facet_wrap(~period)
lvo_period_summary <- ddply(data_nona_vo_voa, c("period"), summarise, N = length(l.vo), mean = mean(l.vo), median = median(l.vo), sd = sd(l.vo))
lvo_period_summary
# Use histogram, the 3 individuals from the Central medeval has no lumbar vo, from the Early medeval there is one individual with 100% lumbar vo and the rest of them at most 25%. Most data are from the Late medeval time, the average is about 8% lumbar vo on average. 


ggplot(data_nona_vo_voa, aes(x = l.vo, col = grave.goods)) + geom_histogram() +
  facet_wrap(~grave.goods)
lvo_grave_summary <- ddply(data_nona_vo_voa, c("grave.goods"), summarise, N = length(l.vo), mean = mean(l.vo), median = median(l.vo), sd = sd(l.vo))
lvo_grave_summary
# Use histogram, people with no grave goods have about 12.5% of lumbar vo on average, people with grave goods have no lumbar vo at all. 
```

### Lumbar VoA Analysis

```{r}
ggplot(data_nona_vo_voa, aes(x = l.voa, col = sex)) + geom_histogram() +
  facet_wrap(~sex)
lvoa_sex_summary <- ddply(data_nona_vo_voa, c("sex"), summarise, N = length(l.voa), mean = mean(l.voa), median = median(l.voa), sd = sd(l.voa))
lvoa_sex_summary
# Females had a slight higher average than males, a lot of females who have lumbar voa have them around 25% - 80%, and males who have lumbar voa have them around 15% - 55%, there is one extreme case of 100%. Note the one who has 100% voa does not 100% vo. 


ggplot(data_nona_vo_voa, aes(x = l.voa, col = age)) + geom_histogram() +
  facet_wrap(~age)
lvoa_age_summary <- ddply(data_nona_vo_voa, c("age"), summarise, N = length(l.voa), mean = mean(l.voa), median = median(l.voa), sd = sd(l.voa))
lvoa_age_summary
# The old age group has the highest lumbar voa, which is around 23%, and those who have lumbar voa had between 35% - 80%. The middle age ground has around 15.6% in average, and those who had it has it between 20% - 65% with 1 extreme cases of 100%. There are around 7 younger individuals that had about 20% of lumbar voa, which is something worth investigating. 


ggplot(data_nona_vo_voa, aes(x = location, y = l.voa, col = location)) + geom_boxplot()
ggplot(data_nona_vo_voa, aes(x = l.voa, col = location)) + geom_histogram() +
  facet_wrap(~location)
lvoa_location_summary <- ddply(data_nona_vo_voa, c("location"), summarise, N = length(l.voa), mean = mean(l.voa), median = median(l.voa), sd = sd(l.voa))
lvoa_location_summary
# There are 4 people buried in church and only 1 of them has no lumbar voa, the rest has around 25% of lumbar voa. Those who were buried in the church yard has about 14% lumbar voa on average, majority of them do not have voa. Those who were buried in Monastic has the expected value of 17.5%, those who have the disease has about 20% - 50% lumbar voa. 


ggplot(data_nona_vo_voa, aes(x = burial, y = l.voa, col = burial)) + geom_boxplot()
ggplot(data_nona_vo_voa, aes(x = l.voa, col = burial)) + geom_histogram() +
  facet_wrap(~burial)
lvoa_burial_summary <- ddply(data_nona_vo_voa, c("burial"), summarise, N = length(l.voa), mean = mean(l.voa), median = median(l.voa), sd = sd(l.voa))
lvoa_burial_summary


ggplot(data_nona_vo_voa, aes(x = period, y = l.voa, col = period)) + geom_boxplot()
ggplot(data_nona_vo_voa, aes(x = l.voa, col = period)) + geom_histogram() +
  facet_wrap(~period)
lvoa_period_summary <- ddply(data_nona_vo_voa, c("period"), summarise, N = length(l.voa), mean = mean(l.voa), median = median(l.voa), sd = sd(l.voa))
lvoa_period_summary


ggplot(data_nona_vo_voa, aes(x = grave.goods, y = l.voa, col = grave.goods)) + geom_boxplot()
ggplot(data_nona_vo_voa, aes(x = l.voa, col = grave.goods)) + geom_histogram() +
  facet_wrap(~grave.goods)
lvoa_grave_summary <- ddply(data_nona_vo_voa, c("grave.goods"), summarise, N = length(l.voa), mean = mean(l.voa), median = median(l.voa), sd = sd(l.voa))
lvoa_grave_summary


```

## Linear Regression

### Between Zp and l.Vo

```{r}
# Deleted excpetion to avoid skewing the data, the ones deleted will be analyzed later
data_no_full <- data_nona_vo_voa%>% filter(l.vo != 100)%>% filter(l.voa != 100)

ggplot(data_no_full, aes(x = l.vo, y = avg_Zp, col = sex)) + geom_point() + geom_smooth(method = "lm") +
  facet_wrap(~sex)

ggplot(data_no_full, aes(x = l.vo, y = avg_Zp, col = age)) + geom_point() + geom_smooth(method = "lm") +
  facet_wrap(~age)
# Not too significant 

ggplot(data_no_full, aes(x = l.vo, y = avg_Zp, col = period)) + geom_point() + geom_smooth(method = "lm") +
  facet_wrap(~period)

ggplot(data_no_full, aes(x = l.vo, y = avg_Zp, col = location)) + geom_point() + geom_smooth(method = "lm") +
  facet_wrap(~location)

ggplot(data_no_full, aes(x = l.vo, y = avg_Zp, col = burial)) + geom_point() + geom_smooth(method = "lm") +
  facet_wrap(~burial)

ggplot(data_no_full, aes(x = l.vo, y = avg_Zp, col = grave.goods)) + geom_point() + geom_smooth(method = "lm") +
  facet_wrap(~grave.goods)

```

### Between Zp and l.VoA

```{r}
ggplot(data_no_full, aes(x = l.voa, y = avg_Zp, col = sex)) + geom_jitter() + geom_smooth(method = "lm") +
  facet_wrap(~sex)

ggplot(data_no_full, aes(x = l.voa, y = avg_Zp, col = age)) + geom_jitter() + geom_smooth(method = "lm") +
  facet_wrap(~age)
# Not too significant 

ggplot(data_no_full, aes(x = l.voa, y = avg_Zp, col = period)) + geom_jitter() + geom_smooth(method = "lm") +
  facet_wrap(~period)

ggplot(data_no_full, aes(x = l.voa, y = avg_Zp, col = location)) + geom_jitter() + geom_smooth(method = "lm") +
  facet_wrap(~location)

ggplot(data_no_full, aes(x = l.voa, y = avg_Zp, col = burial)) + geom_jitter() + geom_smooth(method = "lm") +
  facet_wrap(~burial)

ggplot(data_no_full, aes(x = l.voa, y = avg_Zp, col = grave.goods)) + geom_jitter() + geom_smooth(method = "lm") +
  facet_wrap(~grave.goods)

ggplot(data_no_full, aes(x = l.voa, y=l.vo)) + geom_jitter() + geom_smooth(method = "lm") 
```

## Correlation Coefficient

### Between Zp and l.VoA

```{r}

data_no_full %>%
  filter(period == 'Late')%>%
  select(l.voa, avg_Zp)%>%
  cor()


data_no_full %>%
  filter(sex == 'M')%>%
  select(l.voa, avg_Zp)%>%
  cor()

data_no_full %>%
  filter(sex == 'F')%>%
  select(l.voa, avg_Zp)%>%
  cor()

data_no_full %>%
  filter(age == 'Middle')%>%
  select(l.voa, avg_Zp)%>%
  cor()

data_no_full %>%
  filter(age == 'Old')%>%
  select(l.voa, avg_Zp)%>%
  cor()

data_no_full %>%
  filter(age == 'Young')%>%
  select(l.voa, avg_Zp)%>%
  cor()

data_no_full %>%
  filter(location == 'churchyard')%>%
  select(l.voa, avg_Zp)%>%
  cor()


data_no_full %>%
  filter(burial == 'tomb')%>%
  select(l.voa, avg_Zp)%>%
  cor()

data_no_full %>%
  filter(grave.goods == 'no')%>%
  select(l.voa, avg_Zp)%>%
  cor()


data_no_full %>%
  filter(grave.goods == 'yes')%>%
  select(l.voa, avg_Zp)%>%
  cor()
```

### Between Zp and l.Vo

```{r}
data_no_full %>%
  filter(period == 'Late')%>%
  select(l.vo, avg_Zp)%>%
  cor()


data_no_full %>%
  filter(sex == 'M')%>%
  select(l.vo, avg_Zp)%>%
  cor()

data_no_full %>%
  filter(age == 'Middle')%>%
  select(l.vo, avg_Zp)%>%
  cor()

data_no_full %>%
  filter(location == 'churchyard')%>%
  select(l.vo, avg_Zp)%>%
  cor()


data_no_full %>%
  filter(burial == 'tomb')%>%
  select(l.voa, avg_Zp)%>%
  cor()

data_no_full %>%
  filter(grave.goods == 'no')%>%
  select(l.vo, avg_Zp)%>%
  cor()

data_no_full %>%
  select(l.vo, l.voa)%>%
  cor()


```

Middle age and tomb has the most linear association between L.vo/L.voa and average Zp score.

## Cluster Analysis

We deicded to use hierarchical clustering with complete and average linkage

```{r}
sd.data <- data_nona_vo_voa%>% select(-Zp.left)%>% select(-Zp.right)

hc.complete <- hclust(dist(sd.data), method = "complete")
plot(hc.complete, main = "Complete Linkage")
abline(h = 900, col = "red") # I made the decision to cut it at h = 900 which leaves the dendrogram with 7 clusters.  

hc.average <- hclust(dist(sd.data), method = "average")
plot(hc.average, main = "Average Linkage")
abline(h = 700, col = "red") # I made the decision to cut it at h = 700, whcih leaves the dendrogram with 5 clusters 


```

Adding clusters into the data frame

```{r}
# Now we want to assign each cluster to the main dataframe 

h.cluster.comp <- cutree(hc.complete, 7)
data_nona_vo_voa <- data_nona_vo_voa %>% mutate(h.cluster.comp)
data_nona_vo_voa$h.cluster.comp <- as.factor(data_nona_vo_voa$h.cluster.comp)

h.cluster.avg <- cutree(hc.average, 5)
data_nona_vo_voa <- data_nona_vo_voa %>% mutate(h.cluster.avg)
data_nona_vo_voa$h.cluster.avg <- as.factor(data_nona_vo_voa$h.cluster.avg)


hc.complete <- as.dendrogram(hc.complete)
hc.average <- as.dendrogram(hc.average)

library(dendextend)
dl <- dendlist(highlight_branches(hc.complete), highlight_branches(hc.average))
tanglegram(dl, sort = TRUE, common_subtrees_color_branches = TRUE)
```

```{r}
# the size of each cluster
table(data_nona_vo_voa$h.cluster.comp)

table(data_nona_vo_voa$h.cluster.avg)
```

### Complete Linkage Analysis

```{r}
ggplot(data_nona_vo_voa, aes(x = h.cluster.comp, y= avg_Zp, col = h.cluster.comp)) + geom_boxplot()

cluster.comp.summary.lvoa <- ddply(data_nona_vo_voa, c("h.cluster.comp"), summarise, N = length(l.voa), mean = mean(l.voa), median = median(l.voa), sd = sd(l.voa))
cluster.comp.summary.lvoa

cluster.comp.summary.lvo <- ddply(data_nona_vo_voa, c("h.cluster.comp"), summarise, N = length(l.vo), mean = mean(l.vo), median = median(l.vo), sd = sd(l.vo))
cluster.comp.summary.lvo

cluster.comp.summary.zp <- ddply(data_nona_vo_voa, c("h.cluster.comp"), summarise, N = length(avg_Zp), mean = mean(avg_Zp), median = median(avg_Zp), sd = sd(avg_Zp))
cluster.comp.summary.zp

ggplot(data_nona_vo_voa, aes(x = sex, fill = sex)) + geom_bar() + facet_wrap(~h.cluster.comp)

ggplot(data_nona_vo_voa, aes(x = age, fill = age)) + geom_bar() + facet_wrap(~h.cluster.comp)

ggplot(data_nona_vo_voa, aes(x = period, fill = period)) + geom_bar() + facet_wrap(~h.cluster.comp)

ggplot(data_nona_vo_voa, aes(x = burial, fill = burial)) + geom_bar() + facet_wrap(~h.cluster.comp)

ggplot(data_nona_vo_voa, aes(x = location, fill = location)) + geom_bar() + facet_wrap(~h.cluster.comp)

ggplot(data_nona_vo_voa, aes(x = grave.goods, fill = grave.goods)) + geom_bar() + facet_wrap(~h.cluster.comp)

```

### Average Linkage Analysis

```{r}
ggplot(data_nona_vo_voa, aes(x = h.cluster.avg, y= avg_Zp, col = h.cluster.avg)) + geom_boxplot()

cluster.avg.summary.lvoa <- ddply(data_nona_vo_voa, c("h.cluster.avg"), summarise, N = length(l.voa), mean = mean(l.voa), median = median(l.voa), sd = sd(l.voa))
cluster.avg.summary.lvoa

cluster.avg.summary.lvo <- ddply(data_nona_vo_voa, c("h.cluster.avg"), summarise, N = length(l.vo), mean = mean(l.vo), median = median(l.vo), sd = sd(l.vo))
cluster.avg.summary.lvo

cluster.comp.summary.zp <- ddply(data_nona_vo_voa, c("h.cluster.avg"), summarise, N = length(avg_Zp), mean = mean(avg_Zp), median = median(avg_Zp), sd = sd(avg_Zp))
cluster.comp.summary.zp

ggplot(data_nona_vo_voa, aes(x = sex, fill = sex)) + geom_bar() + facet_wrap(~h.cluster.avg)

ggplot(data_nona_vo_voa, aes(x = age, fill = age)) + geom_bar() + facet_wrap(~h.cluster.avg)

ggplot(data_nona_vo_voa, aes(x = period, fill = period)) + geom_bar() + facet_wrap(~h.cluster.avg)

ggplot(data_nona_vo_voa, aes(x = burial, fill = burial)) + geom_bar() + facet_wrap(~h.cluster.avg)

ggplot(data_nona_vo_voa, aes(x = location, fill = location)) + geom_bar() + facet_wrap(~h.cluster.avg)

ggplot(data_nona_vo_voa, aes(x = grave.goods, fill = grave.goods)) + geom_bar() + facet_wrap(~h.cluster.avg)
```
