---
title: "Clustering Analysis Project 2"
author: "Laura Hu"
date: "2023-04-29"
output:
  pdf_document: default
  html_document: default
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Load library

```{r}
library(ggplot2)
library(ggfortify)
library(GGally)
library(dslabs)
library(caret)
library(modelr)
library(yardstick)
library(dplyr)
library(gridExtra)
library(clustMixType)
```

## Load Data

```{r}
Bully <- read.csv("~/Documents/School /bullying_2018_cleaned.csv")
```

# 

## Initial feature edits to clean up the data

```{r}

#Delete the first row of the data frame and 
Bully <-Bully %>% select(-X) %>% select(-record)
Bully <- Bully[-1,]

#Delete 'Physical_fighting_12_or_more' and 'Physically_attacked_12_or_more' because all variables are false. However, the vairable 'Physically_attacked_12_or_more' shows that there are plenty of students that reported 12.5. We want to change that. 

Bully <-Bully %>% select(-Physical_fighting_12_or_more)

Bully$Physically_attacked_12_or_more <- as.factor(ifelse(Bully$Physically_attacked < 12.1, 'False',
ifelse(Bully$Physically_attacked < 13, 'True')))

sum(is.na(Bully))
```

## Create Partition

```{r}
Bully <-Bully %>% select(-Bullied.outside.school) %>% select(-Cyber.Bullied)

#Create partition (0.7 to 0.3)
set.seed(123)
trainIndex <- createDataPartition(Bully$Bullied.in.School, p = 0.7, 
                                  list = FALSE,
                                  time = 1)

Bully$train <- FALSE 
Bully$train[trainIndex] <- TRUE

Bully_train <- Bully %>% filter(train == TRUE)
Bully_test_1 <- Bully %>%filter(train == FALSE)

```

## Convert Categorical to Numeric for K-Means in the Training Set

```{r}

# Change categorical data to numeric 

#Bullying 
Bully_train$Bullied.in.School <- c(Yes = 1, No = 0)[Bully_train$Bullied.in.School]
#Bully_train$Bullied.outside.school <- c(Yes = 1, No = 0)[Bully_train$Bullied.outside.school]
#Bully_train$Cyber.Bullied <- c(Yes = 1, No = 0)[Bully_train$Cyber.Bullied]

#Sex
Bully_train$Sex <- c(Male = 1, Female = 0)[Bully_train$Sex]

#Felt lonely 
Bully_train$Most_of_the_time_or_always_felt_lonely <- c(Yes = 1, No = 0)[Bully_train$Most_of_the_time_or_always_felt_lonely]

#Missed school or class without permission 
Bully_train$Missed_classes_or_school_without_permission <- c(Yes = 1, No = 0)[Bully_train$Missed_classes_or_school_without_permission]

#Close friends
Bully_train$Close_friends_3_or_more <- c(True = 1, False = 0)[Bully_train$Close_friends_3_or_more]

#Miss school no permission 
Bully_train$Miss_school_no_permission <- c('0 days' = 0, '1 or 2 days' = 1.5, 	
'3 to 5 days' = 4, '6 to 9 days' = 7.5, '10 or more days' = 10 )[Bully_train$Miss_school_no_permission]

#Physcial Attack more than 12 times 
Bully_train$Physically_attacked_12_or_more <- c('True' = 1, 'False' = 0)[Bully_train$Physically_attacked_12_or_more]

```

## Skewness

We ran a series of plots to find out that **'Physically_attacked**', '**Physical_fighting**' are heavily skewed to the right. We need to do some transformations on them to wider the range.

```{r}
# Skew Physcial fighting 
ggplot(Bully_train, aes(x = Physical_fighting)) + geom_density()
Bully_train$Physical_fighting <- log(Bully_train$Physical_fighting + 1) 
ggplot(Bully_train, aes(x = Physical_fighting)) + geom_density()

#Skew Physically attacked 
ggplot(Bully_train, aes(x = Physically_attacked)) + geom_density()
Bully_train$Physically_attacked <- log(Bully_train$Physically_attacked+1)
ggplot(Bully_train, aes(x = Physically_attacked )) + geom_density()
       
```

## Scale data

```{r}
#scale all data
Bully_train_scaled<- Bully_train%>% mutate(across(where(is.numeric), scale))
Bully_train_scaled
```

## Cluster Analysis

```{r}
set.seed(123)
within_ss <- rep(NA, 15)
for (i in 1:15){
  km_res <- kmeans(x = Bully_train_scaled,
                  center = i,
                  nstart = 5)
  within_ss[i] <- km_res$tot.withinss
}

plot(1:15, within_ss, type = "b", ylab = "Objective Function", xlab = "# Clusters",
main = "Scree Plot")

```

From the graph, we can see that two obvious choices are 9 and 11, we want to evaluate both choices

### Cluster 9

```{r}
res_9 <- kmeans(Bully_train_scaled,
                   centers = 9,
                   nstart = 5)

#Add the clusters into the original data frame
result <- res_9$cluster
result <- as.factor(result)
size <- res_9$size
size

#New data frame that included the clustering results 
Bully_trained_cluster_9 <- data.frame(Bully_train, result)

#Visualize the clusters 
ggplot(Bully_trained_cluster_9, aes(x = result, fill = factor(Bullied.in.School))) + geom_bar(position = "stack") + labs(fill = "Bullied in School") + scale_fill_manual(values = c("#DADAEB", "#9E9AC8"))

View(Bully_trained_cluster_9)

```

### Cluster 13

```{r}
res_13 <- kmeans(Bully_train_scaled,
                   centers = 11,
                   nstart = 5)

#Add the clusters into the original data frame
result_13 <- res_13$cluster
result_13 <- as.factor(result_13)
size <- res_13$size
size

#New data frame that included the clustering results 
Bully_trained_cluster_13 <- data.frame(Bully_train, result_13)

#Visualize the clusters 
ggplot(Bully_trained_cluster_13, aes(x = result_13, fill = factor(Bullied.in.School))) + geom_bar(position = "stack") + labs(fill = "Bullied in School") + scale_fill_manual(values = c("#DADAEB", "#9E9AC8"))

```

We choose **9 clusters** for our prediction because it's easier, additional clusters only reinforces the information that can be found in cluster 9 which can overcomplicate the classification process.

From the graph above, we can see that **cluster 1** are all bullied in school, **cluster 3,4,5** are not at all bullied in school (there's a small portion of cluster 4 that are bullied, but that does not really affect the classification process by much)

We want to focus more on [cluster 2,6,7,8 and 9]{.underline}. Cluster 2 and 9 only has a small portion that are bullied, cluster 7,8 has a little bit less than half, cluster 6 has just about half.

## Visualization of the clusters

To further investigate the characteristics of bullying, we want to visualize the each clusters.

( Because most of the data points are discrete, it is hard to visualize. I tried using bubble chart, but I cannot find any good indicator as to why those groups were bullied. )

### Physically Attacked

```{r}
Bully_trained_cluster_9$Bullied.in.School <- as.factor(Bully_trained_cluster_9$Bullied.in.School)

ggplot(Bully_trained_cluster_9, aes(x = result, fill = factor(Physically_attacked))) + geom_bar(position = "stack") + labs(fill = "Physically_attacked") 

ggplot(Bully_trained_cluster_9, aes(x = result, y = Custom_Age, size = factor(Physically_attacked), col = as.factor(Bullied.in.School))) + geom_point(alpha = 0.3) + scale_size_manual(values = c(0.001 , 1, 3, 5, 7, 9, 11, 13))

```

### Physical fighting

```{r}
Bully_trained_cluster_9$Bullied.in.School <- as.factor(Bully_trained_cluster_9$Bullied.in.School)

ggplot(Bully_trained_cluster_9, aes(x = result, fill = factor(Physical_fighting))) + geom_bar(position = "stack") + labs(fill = "Physical fighting") 

ggplot(Bully_trained_cluster_9, aes(x = result, y = Custom_Age, size = factor(Physical_fighting), col = as.factor(Bullied.in.School))) + geom_point(alpha = 0.3) + scale_size_manual(values = c(0.001 , 1, 3, 5, 7, 9, 11, 13))

ggplot(Bully_trained_cluster_9, aes(x = result, y = Bully_trained_cluster_9$Physical_fighting, fill = result)) + geom_boxplot()
```

```{r}

ggplot(Bully_trained_cluster_9, aes(x = Felt_lonely, fill = factor(Bullied.in.School))) + geom_bar(position = "stack") + labs(fill = "Bullied.in.School") 

ggplot(Bully_trained_cluster_9, aes(x = result, y = Custom_Age, size = factor(Physical_fighting), col = as.factor(Bullied.in.School))) + geom_point(alpha = 0.3) + scale_size_manual(values = c(0.001 , 1, 3, 5, 7, 9, 11, 13))
```

```{r}
ggplot(Bully_trained_cluster_9, aes(x = result, fill = factor(Felt_lonely))) + geom_bar(position = "stack") + labs(fill = "Felt_lonely") 

ggplot(Bully_trained_cluster_9, aes(x = result, y = Felt_lonely, col = as.factor(Bullied.in.School))) + geom_point(alpha = 0.3)
```
