---
title: "STA4990 Titanic"
author: "Laura Hu"
date: "2023-03-29"
output:
  pdf_document: default
  html_document: default
---

```{r}
#Load related packages 
library(ggplot2)
library(ggfortify)
library(GGally)
library(dslabs)
library(caret)
library(modelr)
library(yardstick)
library(dplyr)
library(gridExtra)
```

```{r}
ship <- read.csv("~/Documents/School /titanic/ship_train.csv")
```

## Initial feature engineering to clean up the raw data

### Name

-   First regarding the name title, we found that '**Master**' and '**Miss**' refers to kids below the age 12

-   After the age of 12, most people are referred as 'Miss' '**Mrs**' and '**Mr**'

-   We found that there are also titles like '**Rev**'

-   There are other titles like 'Lady', 'Sir' and 'Countess' and more, we are going to name them all '**Others**'

```{r}

for(i in 1:nrow(ship)){
  name <- ship$Name[i]
  has_master <- grepl("Master", name)
  has_rev <- grepl("Rev.", name)
  has_miss <- grepl("Miss.", name)
  has_mrs <- grepl("Mrs.", name)
  has_mr <- grepl("Mr.", name)
  
  
  if(has_master == TRUE){
    ship$Name[i] <- 'Master' 
  } else if(has_rev == TRUE){
    ship$Name[i] <- 'Rev'
  } else if (has_miss == TRUE){
    ship$Name[i] <- 'Miss'
  } else if (has_mrs == TRUE){
    ship$Name[i] <- 'Mrs'
  } else if (has_mr == TRUE){
    ship$Name[i] <- 'Mr'
  } else{
    ship$Name[i] <- 'Other'
  }
}
```

### Age

There are a lot of 'Age' missing from the data set, what we want to do is fill in all the NA values with the average age in the data set

```{r}
ship$Age[is.na(ship$Age)] <- mean(ship$Age, na.rm = TRUE)
```

### Ticket

-   Variable names: WEP, W./C, STON/O2, SOTON/OQ, SC/AH, SOC, S.O./P.P., SC/, PC, LINE, F.C.C, CA, C, A5, 1,2,3,4,5,6,7,8,9

-   Singles: SW/PP 751,PP 4348, S.P. 3464, SO/C 14885

I didn't do any feature engineering on this one, I didn't really bother.

## Remove Variables

```{r}
ship <- ship %>% 
  select(-Ticket)%>% 
  select(-Cabin)%>% 
  select(-PassengerId) %>% 
  mutate(Survived = (Survived > 0))
```

## Initial Observation

```{r}
#ggpairs(ship, aes(color = Survived))
```

### 'Age'

```{r}
ggplot(ship, aes(x = Age, y = Name, color = Survived)) + geom_jitter(alpha = 0.5)
```

The age cutoff for 'Master' is around 12, we should probably do further investigation on the variable 'Miss'

### 'Name'

```{r}
ggplot(ship, aes(x = Name, y = Pclass, color = Survived)) + geom_jitter(alpha = 0.5)
```

```{r}
ggplot(ship, aes(x = Name, y = SibSp, color = Survived)) + geom_jitter(alpha = 0.5)
```

```{r}
ggplot(ship, aes(x = Name, y = Parch, color = Survived)) + geom_jitter(alpha = 0.5)
```

```{r}
ggplot(ship, aes(x = Pclass, y = SibSp, color = Survived)) + geom_jitter(alpha = 0.5)
```

```{r}
ggplot(ship, aes(x = Pclass, y = Parch, color = Survived)) + geom_jitter(alpha = 0.5)
```

'Master', 'Miss' and 'Mrs' in class 1,2 with less than 3 siblings & Parch are more likely to survive.

Additionally We can see that almost all class 1,2 didn't bring more than 2 siblings. So Class 1,2 implies SibSp \< 3 & Parch \< 3

### 'Fare'

```{r}
ggplot(ship, aes(x = Pclass, y = Fare, color = Survived)) + geom_jitter(alpha = 0.5)
```

```{r}
ggplot(ship, aes(x = Age, y = Fare, color = Survived)) + geom_jitter(alpha = 0.5)
```

```{r}
ggplot(ship, aes(x = SibSp, y = Fare, color = Survived)) + geom_jitter(alpha = 0.5)
```

People that are already in class 1 that paid more than 50 are more likely to survive.

## 

## Additional Feature Engineering

## 'Companion'

```{r}
ship$Companion <- ship$SibSp + ship$Parch
```

## 'Rich' 

Add variables called 'Rich' for people that are in class 1 and paid more than \$50

Add variable called 'Prime' for 'Masters', 'Miss' and 'Mrs' in class 1,2

```{r}
ship$Rich <- as.factor(ifelse(ship$Pclass < 2 & ship$Fare < 30, 0,
                              ifelse(ship$Pclass < 2 & ship$Fare< 600, 1,
                                     ifelse(ship$Pclass < 4 & ship$Fare < 600, 0)))) 
```

### Checking to see if it is significant

```{r}
ggplot(ship, aes(x = Rich, y = Fare, color = Survived)) + geom_jitter(alpha = 0.5)

```

```{r}
ggplot(ship, aes(x = Rich, y = Pclass, color = Survived)) + geom_jitter(alpha = 0.5)
```

## 'Prime'

'Master', 'Miss' and 'Mrs' in class 1,2 with less than 3 siblings & Parch are more likely to survive.

```{r}
#First add a new varaible that converts all the names into numeric titles 

ship$Name_num <- c(Master = 0, Miss = 0, Mrs = 0, Mr = 1, Rev = 1, Other = 1)[ship$Name]

#Then Use the new Variable to create the varaible 'Prime'
ship$Prime <- as.factor(ifelse(ship$Name_num < 1 & ship$Pclass < 3, 1,
                       ifelse(ship$Name_num < 2 & ship$Pclass < 4, 0)))

```

### Graph to check significance 

```{r}
ggplot(ship, aes(x = Prime, y = Name_num, color = Survived)) + geom_jitter(alpha = 0.5)
```

## Split data into training set and test set

```{r}
#Create partition (0.7 to 0.3)
set.seed(165)
trainIndex <- createDataPartition(ship$Survived, p = 0.7, 
                                  list = FALSE,
                                  time = 1)

ship$train <- FALSE 
ship$train[trainIndex] <- TRUE

ship_train <- ship %>% filter(train == TRUE)
ship_test_1 <- ship%>%filter(train == FALSE)
```

## 

## Fit Models 

### QDA

```{r}
fit_qda <- train(as.factor(Survived) ~ Prime + Rich + Age + Companion,
                 data = ship_train,
                 method = "qda",
                 trace = FALSE)

```

### LDA

```{r}
fit_lda <- train(as.factor(Survived) ~ Prime + Rich + Age + Companion,
                 data = ship_train,
                 method = "lda",
                 trace = FALSE)
```

### KNN

```{r}
fit_knn <- train(as.factor(Survived) ~ Prime + Rich + Age + Companion,
                 data = ship_train,
                 method = "knn")
```

### Naive Bayes

```{r}
fit_nb <- train(as.factor(Survived) ~ Prime + Rich + Age + Companion,
                 data = ship_train,
                 method = "naive_bayes")
```

### Logistic Regression 

```{r}
fit_log <- train(as.factor(Survived) ~ Prime + Rich + Age + Companion,
                 data = ship_train,
                 method = "glmnet",
                 family = "multinomial",
                 trace = FALSE)
```

## Checking accuracy by fitting into the test set 

```{r}
ship_test_1 <- ship_test_1 %>%
  mutate(Survived_qda = predict(fit_qda, newdata = ., type = "raw"))%>%
  mutate(qda_prob = predict(fit_qda, newdata = ., type = "prob")$'TRUE')
```

```{r}
ship_test_1 <- ship_test_1 %>%
  mutate(Survived_lda = predict(fit_lda, newdata = ., type = "raw"))%>%
  mutate(lda_prob = predict(fit_lda, newdata = ., type = "prob")$'TRUE')
```

```{r}
ship_test_1 <- ship_test_1 %>%
  mutate(Survived_knn = predict(fit_knn, newdata = ., type = "raw"))%>%
  mutate(knn_prob = predict(fit_qda, newdata = ., type = "prob")$'TRUE')
```

```{r}
ship_test_1 <- ship_test_1 %>%
  mutate(Survived_nb = predict(fit_nb, newdata = ., type = "raw"))%>%
  mutate(nb_prob = predict(fit_nb, newdata = ., type = "prob")$'TRUE')
```

```{r}
ship_test_1 <- ship_test_1 %>%
  mutate(Survived_log = predict(fit_log, newdata = ., type = "raw"))
#%>%mutate(log_prob = predict(fit_log, newdata = ., type = "prob")$'TRUE')
```

Accuracy

```{r}
ship_test_1 %>% accuracy(truth = as.factor(Survived),
                       estimate = Survived_qda)
```

```{r}
ship_test_1 %>% accuracy(truth = as.factor(Survived),
                       estimate = Survived_lda)
```

```{r}
ship_test_1 %>% accuracy(truth = as.factor(Survived),
                       estimate = Survived_knn)
```

```{r}
ship_test_1 %>% accuracy(truth = as.factor(Survived),
                       estimate = Survived_nb)
```

```{r}
ship_test_1 %>% accuracy(truth = as.factor(Survived),
                       estimate = Survived_log)
```

## Fit to Test Set 

```{r}
ship_test <- read.csv("~/Documents/School /titanic/ship_test.csv")
```

```{r}
for(i in 1:nrow(ship_test)){
  name <- ship_test$Name[i]
  has_master <- grepl("Master", name)
  has_rev <- grepl("Rev.", name)
  has_miss <- grepl("Miss.", name)
  has_mrs <- grepl("Mrs.", name)
  has_mr <- grepl("Mr.", name)
  
  
  if(has_master == TRUE){
    ship_test$Name[i] <- 'Master' 
  } else if(has_rev == TRUE){
    ship_test$Name[i] <- 'Rev'
  } else if (has_miss == TRUE){
    ship_test$Name[i] <- 'Miss'
  } else if (has_mrs == TRUE){
    ship_test$Name[i] <- 'Mrs'
  } else if (has_mr == TRUE){
    ship_test$Name[i] <- 'Mr'
  } else{
    ship_test$Name[i] <- 'Other'
  }
}
```

```{r}
ship_test$Age[is.na(ship_test$Age)] <- mean(ship_test$Age, na.rm = TRUE)

ship_test$Fare[is.na(ship_test$Fare)] <- mean(ship_test$Fare, na.rm = TRUE)
```

```{r}
ship_test$Companion <- ship_test$SibSp + ship_test$Parch
```

```{r}
ship_test$Rich <- as.factor(ifelse(ship_test$Pclass < 2 & ship_test$Fare < 30, 0,
                              ifelse(ship_test$Pclass < 2 & ship_test$Fare< 600, 1,
                                     ifelse(ship_test$Pclass < 4 & ship_test$Fare < 600, 0)))) 
```

```{r}
#First add a new varaible that converts all the names into numeric titles 

ship_test$Name_num <- c(Master = 0, Miss = 0, Mrs = 0, Mr = 1, Rev = 1, Other = 1)[ship_test$Name]

#Then Use the new Variable to create the varaible 'Prime'
ship_test$Prime <- as.factor(ifelse(ship_test$Name_num < 1 & ship_test$Pclass < 3, 1,
                       ifelse(ship_test$Name_num < 2 & ship_test$Pclass < 4, 0)))
```

```{r}
ship_test <- ship_test %>%
  mutate(Survived_log = predict(fit_log, newdata = ., type = "raw"))
```

```{r}
ship_test <- ship_test %>% 
  mutate(Survived = (Survived_log == TRUE))%>%
  select(-Ticket)%>% 
  select(-Cabin)%>% 
  select(-Pclass)%>%
  select(-Name)%>%
  select(-Sex)%>%
  select(-Age)%>%
  select(-SibSp)%>%
  select(-Parch)%>%
  select(-Fare)%>%
  select(-Embarked)%>%
  select(-Companion)%>%
  select(-Rich)%>%
  select(-Prime)%>%
  select(-Name_num)
```

```{r}
ship_test$Survived <- as.integer(as.logical(ship_test$Survived))
ship_test <- ship_test %>%
  select(-Survived_log)
```

```{r}
write.csv(ship_test, "~/Documents/School /titanic/titanic_pred_1.csv", row.names=FALSE)
```
